Supplement to Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-03-24 trunk/paper/supplement.tex

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\renewcommand{\cite}{\citep}

\begin{document}
\firstpage{1}

\title[Physiology phenotypes]{Semantic integration of physiology
  phenotypes with an application to the Cellular Phenotype Ontology:
  Supplement 1}

\author[Hoehndorf \textit{et~al}]{Robert Hoehndorf$^{1}$\footnote{to
    whom correspondence should be addressed}, Midori
  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and
  Georgios V. Gkoutos$^{1}$}

\address{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
  Court Road, Cambridge CB2 1GA, UK\\
  $^{3}$Institute for Medical Informatics, Statistics and
  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
  Leipzig, Germany\\
  $^{4}$European Bioinformatics Institute, Wellcome Trust Genome
  Campus, Hinxton, Cambridge, Cambridge CB10 1SD, UK}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

@@

\subsection{Detailed axiom patterns}

Phenotype of $X$
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (part-of some X))
\end{verbatim}

Abnormality of $X$
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (part-of some (X and has-quality some Quality)))
\end{verbatim}

Abnormal $X$ morphology:
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (X and has-quality some Morphology))
\end{verbatim}

@@

>>>>

\documentclass{article}
\begin{document}

\title{Supplement to Semantic integration of physiology phenotypes with an
  application to the Cellular Phenotype Ontology}

\author{Robert Hoehndorf$^{1}$\footnote{to whom correspondence should
    be addressed}, Midori A. Harris$^2$, Heinrich Herre$^3$, \\Gabriella
  Rustici$^4$ and Georgios V. Gkoutos$^{1}$}

\date{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
  Court Road, Cambridge CB2 1GA, UK\\
  $^{3}$Institute for Medical Informatics, Statistics and
  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
  Leipzig, Germany\\
  $^{4}$European Bioinformatics Institute, Wellcome Trust Genome
  Campus, Hinxton, Cambridge, Cambridge CB10 1SD, UK}

@@

\subsection{Detailed axiom patterns}
Here, we present the precise formulation of axiom patterns. We use $X$
and $A$ as variables in these patterns that are replaced with specific
classes from the GO or ChEBI ontologies.

\paragraph{Phenotype of X:}
If $X$ is a material object, we assert:
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (part-of some X))
\end{verbatim}

If $X$ is a cellular biological process, we assert:
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (participates-in some (part-of some X)))
\end{verbatim}

\paragraph{Abnormality of X:}
If $X$ is a material object:
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (part-of some (X and has-quality some Quality)))
\end{verbatim}

If $X$ is a process, we create a class that does not have any formal
constraints. In CPO, abnormalities of $X$ processes include both
abnormalities of $X$ and abnormality of regulation of $X$. A
formulation of this constraint requires the use of disjunction. Since
we limit ourselves to the use of OWL EL and disjunction is not a valid
logical operator in OWL EL, we do not add any axiom to CPO in this case.

\paragraph{Abnormal morphology and physiology, and absence:}
Abnormal $X$ morphology:
\begin{verbatim}
'X phenotype' EquivalentTo:
  phenotype-of some (has-part some
    (X and has-quality some Morphology))
\end{verbatim}

@@

Formalizing absence of $X$ requires the use of negation. Since
negation is not allowed in OWL EL, we create a class for absence that
we directly assert to be a subclass of {\em Abnormality of X}, without
adding further axioms.

\paragraph{Single occurrence abnormalities:}
Abnormality of single occurrence of $X$:
\begin{verbatim}
'Abnormality of single occurrence of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (part-of some X and
has-quality some Quality)))
\end{verbatim}

Increased duration of X:
\begin{verbatim}
'Increased duration of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (X and
has-quality some 'Increased duration')))
\end{verbatim}

Decreased duration of X:
\begin{verbatim}
'Decreased duration of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (X and
has-quality some 'Decreased duration')))
\end{verbatim}

\paragraph{Abnormalities of X regulation:}
Abnormality of regulation of $X$:
\begin{verbatim}
'Abnormality of regulation of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (regulates some X and
has-quality some (Quality and towards some X))))
\end{verbatim}

Increased frequency of $X$:
\begin{verbatim}
'Increased frequency of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (regulates some X and
has-quality some ('Increased frequency' and towards some X))))
\end{verbatim}

Decreased frequency of $X$:
\begin{verbatim}
'Decreased frequency of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (regulates some X and
has-quality some ('Decreased frequency' and towards some X))))
\end{verbatim}

Late onset of $X$:
\begin{verbatim}
'Late onset of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (regulates some X and
has-quality some (Delayed and towards some X))))
\end{verbatim}

Early onset of $X$:
\begin{verbatim}
'Early onset of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (regulates some X and
has-quality some (Premature and towards some X))))
\end{verbatim}

\paragraph{Abnormality of inputs and outputs (rates):}
If, by virtue of GO-XP definitions, we are able to identify inputs and
outputs of $X$, we define the following classes:

Increased mass of $A$ as output in regulation of $X$:
\begin{verbatim}
'Decreased mass of A as output in regulation of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (regulates some X and
has-output some (A and has-quality some 'Increased mass'))))
\end{verbatim}

Decreased mass of $A$ as output in regulation of $X$:
\begin{verbatim}
'Decreased mass of A as output in regulation of X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (regulates some X and
has-output some (A and has-quality some 'Decreased mass'))))
\end{verbatim}

Increased mass of $A$ as output in $X$:
\begin{verbatim}
'Increased mass of A as output in X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (X and
has-output some (A and has-quality some 'Increased mass'))))
\end{verbatim}

Decreased mass of $A$ as output in $X$:
\begin{verbatim}
'Decreased mass of A as output in X' EquivalentTo:
phenotype-of some (has-part some (
participates-in some (X and
has-output some (A and has-quality some 'Decreased mass'))))
\end{verbatim}

We assert similar axioms for the inputs of $X$.
